A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-08-10 doc/jcomp2012/jcomp2012.tex
final edits for jcomp draft

In previous work~\cite{evans_2003}, we developed a residual Monte
Carlo method that was an application of Halton's Sequential Monte
Carlo method~\cite{halton_1962,halton_1994} to the one-dimensional
equilibrium (1T) thermal radiation diffusion equation.  At that time,
Halton's method was not widely known in computational physics.  While
extending our method to three dimensions, we have discovered that the
Sequential Monte Carlo method is actually a variant of iterative
refinement cast as a residual method.  Using Monte Carlo to accelerate
a fixed-point (Richardson) iteration, our new method surpasses our
previous one dimensional method and allows for extension to more
general solution techniques.

@@

In order to address the convergence issues plaguing Monte Carlo
solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
Monte Carlo method.  This algorithm demonstrated dramatically improved
convergence over regular Monte Carlo.  Nonetheless, Halton's method
has not gained widespread use in computational physics.  In the next
sections, we will briefly describe the Monte Carlo methods that can be
used to solve linear systems and expand them to present our new
synthetic acceleration method.

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permutations were computed per
MCSA iteration with a weight cutoff of $1.0E-4$ for the stochastic
linear solver and a convergence tolerance of $1.0E-8$ for the MCSA
iterative solver. All computations presented in this section were
completed on a 2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3
memory. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for
each case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size.

@@

where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
$[\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
heat capacity of the material, $\rho\equiv\rho(\ve{r})$
$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material,
$a=0.01372$ $[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
the radiation constant, $c=299.79$ $[\text{cm}\cdot\text{sh}^{-1}]$ is
the vacuum light speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is
the temperature that characterizes both the radiation and the
material.  The opacity, $\ros\equiv\ros$(\ve{r},T(\ve{r},t))
$[\text{cm}^{-1}]$, is the Rosseland Mean opacity.  The source,
$Q\equiv Q(\ve{r},t)$
$[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{sh}^{-1}]$, is assumed to
be independent of $T$.

We choose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{Jerks}\cdot\text{cm}^{-2}\cdot\text{sh}^{-1}]$, as the primary
state variable.  The intensity is defined as:

@@

The second item is quite important for Monte Carlo solutions of linear
systems, and this type of scaling should be performed regardless of
any additional preconditioner choices.

A side effect of this preconditioner choice is that the preconditioned
system, $\ve{M}^{-1}\ve{D}$, is no longer SPD.  This situation can be
rectified by doing right-left preconditioning and defining

\begin{equation}
  \ve{M} = \ve{L}\ve{L}^{T}\:,
\end{equation}

where $\ve{L} = \ve{L}^{T} = \sqrt{\ve{M}}$.  Applying right-left
preconditioning gives

\begin{equation}
  \begin{gathered}
    \ve{M}^{-1}\ve{D}\ve{M}^{-1}\ve{M}\bphi =
    \ve{M}^{-1}\ve{q}\:,\\ (\ve{LL}^T)^{-1}\ve{D}(\ve{LL}^T)^{-1}\ve{LL}^T\bphi
    = (\ve{LL}^T)^{-1}\ve{q}\:,
    \\ \ve{L}^{-T}\ve{L}^{-1}\ve{D}\ve{L}^{-T}\ve{L}^T\phi =
    \ve{L}^{-T}\ve{L}^{-1}\ve{q}\:,\\ (\ve{L}^{-1}\ve{D}\ve{L}^{-T})\ve{L}^T\bphi
    = \ve{L}^{-1}\ve{q}\:,\\ \hat{\ve{D}}\hat{\bphi} = \hat{\ve{q}}\:.
  \end{gathered}
  \label{eq:right-left-preconditioning}
\end{equation}

This type of preconditioning is also referred to as row-column
scaling.  The resulting preconditioned matrix, $\hat{\ve{D}}$, is now
SPD, and we can use Conjugate Gradient (CG) solvers.  Therefore, we
will use this preconditioning scheme for deterministic solutions of
Eq.~(\ref{eq:diffusion-matrix}).  However, given that the convergence
properties of $\hat{\ve{D}}$ are similar to $\ve{M}^{-1}\ve{D}$ and
the Monte Carlo solvers do not require SPD matrices, we will be
content with solving Eq.~(\ref{eq:preconditioned-diffusion}) using
MCSA.

The preconditioned system we will solve is defined in
Eq.~(\ref{eq:preconditioned-diffusion}).  Applying $\ve{M}^{-1}$ to
Eq.~(\ref{eq:discrete-diffusion}) yields

@@

Here we examine the effectiveness of the MCSA method and compare it to
traditional deterministic solution methods.  The deterministic methods
considered are preconditioned Conjugate Gradient (PCG) and
preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
method uses right-left preconditioning a shown in
Eq.~(\ref{eq:right-left-preconditioning}).  The PFIX scheme is used to
solve Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to
the system being solved with MCSA.  We examine each method on two 3D
problems: a Marshak wave problem and a multi-material duct problem.

@@

The Marshak wave problem under consideration in this study has an
incoming 1~keV blackbody boundary flux at $x=0$ and propagates in the
$x$-dimension in time.  In other words, this a 3D representation of a
1D problem.  The full set of problem conditions is:

\begin{center}
  \begin{tabular}{ll}\hline
    B.C. & $f_b^{+}(x=0,T=1.0) = \frac{ac}{4}T^4$ \\ & $\ve{F} =
    0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+}, z^{-}, z^{+}\}$
    \\ material properties & $\rho =
    3.0\,\text{g}\cdot\text{cm}^{-3}$ \\ & $\Cv =
    0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
    $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
    100T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ \\  mesh & $\Di = \Dj
    = \Dk = 0.01\,\text{cm}$ \\ & $N_i = 100\quad N_j = N_k = 4$
    \\ \hline
  \end{tabular}
\end{center}

The problem is run to $10$~shakes.

@@

\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=10$~sh.  The
    analytic solution is compared to the MCSA solution.  Note that the
    MCSA, PCG, and PFIXED methods all give numerically identical
    answers.}
  \label{fig:marshak_10}
\end{figure}

@@

\begin{center}
  \begin{tabular}{ll}\hline
    B.C. & $f_b^{+}(x=0,T=0.5) = \frac{ac}{4}T^4$\\ & $\ve{F} =
    0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+}, z^{-}, z^{+}\}$
    \\  material properties & $\rho =
    1.5\,\text{g}\cdot\text{cm}^{-3}$ (duct) \\ & $\rho =
    8.0\,\text{g}\cdot\text{cm}^{-3}$ (wall) \\ & $\rho =
    4.0\,\text{g}\cdot\text{cm}^{-3}$ (foil) \\ & $\Cv =
    0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
    $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
    T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ (duct) \\ & $\sigma =
    1000T^{-1}\,\text{cm}^2\cdot\text{g}^{-1}$ (wall) \\ & $\sigma =
    5T^{-2}\,\text{cm}^2\cdot\text{g}^{-1}$ (foil) \\  mesh & $\Di =
    \Dj = \Dk = 0.025\,\text{cm}$ \\ & $N_i = N_j = N_k = 60$
    \\ \hline
  \end{tabular}
\end{center}

A 2D plot of the temperature at 100~sh elapsed time is illustrated in
Fig.~\ref{fig:multi-mat_T}.  The time-evolution of the temperature is
shown at 4 edit points in Fig.~\ref{fig:edit-points}.  All three
methods, PCG, PFIX, and MCSA, give numerically identical answers when
using a stopping criterion of $\sn{1}{-8}$.

@@

\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
  \caption{ Two-dimensional plot of the temperature at 100~sh on a
    cut-plane positioned at the mid-point of the $z$-axis.  The
    positions of the edit points where the time-evolution of the
    temperature is plotted in Fig.~\ref{fig:edit-points} are indicated
    by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

@@

\begin{table}[htbp!]
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
      18 & 1.03 \\  PFIX & 38 & 1.43 \\  MCSA & 20 & 1.0
      \\ \hline\hline
    \end{tabular}
  \end{center}
  \caption{ Comparison of solution methods for the multi-material
    problem. All methods were run with a stopping criterion of
    $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
    \sn{1}{-3}$.  The problem was run to an elapsed time of 10~sh.
    The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}
  \label{tab:multimat_comparison}
\end{table}

>>>>

In previous work~\cite{evans_2003}, we developed a residual Monte
Carlo method that was an application of Halton's Sequential Monte
Carlo method~\cite{halton_1962,halton_1994} to the one-dimensional
equilibrium (1T) thermal radiation diffusion equation.  At that time,
Halton's method was not widely known in the general computational
physics community.  While extending our method to three dimensions, we
have realized that the Sequential Monte Carlo method is actually a
variant of iterative refinement cast as a residual method. We have
developed a new method in which Monte Carlo is used to accelerate a
fixed-point (Richardson) iteration. This algorithm surpasses our
previous one dimensional method and allows for extension to more
general solution techniques.

@@

In order to address the convergence issues plaguing Monte Carlo
solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
Monte Carlo method.  This algorithm demonstrated dramatically improved
convergence over regular Monte Carlo.  Nonetheless, Halton's method
has not gained widespread use in computational physics
applications. In the next sections, we will briefly describe the Monte
Carlo methods that can be used to solve linear systems and expand them
to present our new synthetic acceleration method.

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permutations were computed per
MCSA iteration with a weight cutoff of \sn{1}{-4} for the stochastic
linear solver and a convergence tolerance of \sn{1}{-8} for the MCSA
iterative solver. All computations presented in this section were
completed on a 2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3
memory. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for
each case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size.

@@

where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
$[\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
heat capacity of the material, $\rho\equiv\rho(\ve{r})$
$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material,
$a=0.01372$ $[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
the radiation constant, $c=29.979$ $[\text{cm}\cdot\text{ns}^{-1}]$ is
the vacuum light speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is
the temperature that characterizes both the radiation and the
material.  The opacity, $\ros\equiv\ros$(\ve{r},T(\ve{r},t))
$[\text{cm}^{-1}]$, is the Rosseland Mean opacity.  The source,
$Q\equiv Q(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{ns}^{-1}]$, is assumed to
be independent of $T$.

We choose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary
state variable.  The intensity is defined as:

@@

The second item is quite important for Monte Carlo solutions of linear
systems, and this type of scaling should be performed regardless of
any additional preconditioner choices. We note that the resulting
system does not have to be SPD for the MCSA algorithm, thus any
suitable choice of $\ve{M}$ will suffice.  The preconditioned system
we will solve is defined in Eq.~(\ref{eq:preconditioned-diffusion}).
Applying $\ve{M}^{-1}$ to Eq.~(\ref{eq:discrete-diffusion}) yields

@@

Here we examine the effectiveness of the MCSA method and compare it to
traditional deterministic solution methods.  The deterministic methods
considered are preconditioned Conjugate Gradient (PCG) and
preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
method uses right-left preconditioning with
$\ve{M}=\ve{L}\ve{L}^{-1}$.  The PFIX scheme is used to solve
Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
system being solved with MCSA.  We examine each method on two 3D
problems: a Marshak wave problem and a multi-material duct problem.

@@

The Marshak wave problem under consideration in this study has an
incoming 1~keV blackbody boundary flux at $x=0$ and propagates in the
$x$-dimension in time as outlined in Ref.~\cite{larsen_1980}.  In
other words, this a 3D representation of a 1D problem.  The full set
of problem conditions is:

\begin{center}
  \begin{tabular}{ll}\hline
    B.C. & $f_b^{+}(x=0,T=1.0) = \frac{ac}{4}T^4$ \\ & $\ve{F} =
    0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+}, z^{-}, z^{+}\}$
    \\ material properties & $\rho =
    3.0\,\text{g}\cdot\text{cm}^{-3}$ \\ & $\Cv =
    0.1\,\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
    $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
    100T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ \\  mesh & $\Di = \Dj
    = \Dk = 0.01\,\text{cm}$ \\ & $N_i = 100\quad N_j = N_k = 4$
    \\ \hline
  \end{tabular}
\end{center}

The problem is run to $100$~nanoseconds.

@@

\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The
    analytic solution is compared to the MCSA solution.  Note that the
    MCSA, PCG, and PFIXED methods all give numerically identical
    answers.}
  \label{fig:marshak_10}
\end{figure}

@@

\begin{center}
  \begin{tabular}{ll}\hline
    B.C. & $f_b^{+}(x=0,T=0.5) = \frac{ac}{4}T^4$\\ & $\ve{F} =
    0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+}, z^{-}, z^{+}\}$
    \\  material properties & $\rho =
    1.5\,\text{g}\cdot\text{cm}^{-3}$ (duct) \\ & $\rho =
    8.0\,\text{g}\cdot\text{cm}^{-3}$ (wall) \\ & $\rho =
    4.0\,\text{g}\cdot\text{cm}^{-3}$ (foil) \\ & $\Cv =
    0.1\,\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
    $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
    T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ (duct) \\ & $\sigma =
    1000T^{-1}\,\text{cm}^2\cdot\text{g}^{-1}$ (wall) \\ & $\sigma =
    5T^{-2}\,\text{cm}^2\cdot\text{g}^{-1}$ (foil) \\  mesh & $\Di =
    \Dj = \Dk = 0.025\,\text{cm}$ \\ & $N_i = N_j = N_k = 60$
    \\ \hline
  \end{tabular}
\end{center}

A 2D plot of the temperature at 1000~ns elapsed time is illustrated in
Fig.~\ref{fig:multi-mat_T}.  The time-evolution of the temperature is
shown at 4 edit points in Fig.~\ref{fig:edit-points}.  All three
methods, PCG, PFIX, and MCSA, give numerically identical answers when
using a stopping criterion of $\sn{1}{-8}$.

@@

\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
  \caption{ Two-dimensional plot of the temperature at 1000~ns on a
    cut-plane positioned at the mid-point of the $z$-axis.  The
    positions of the edit points where the time-evolution of the
    temperature is plotted in Fig.~\ref{fig:edit-points} are indicated
    by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

@@

\begin{table}[htbp!]
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
      18 & 1.03 \\  PFIX & 38 & 1.43 \\  MCSA & 20 & 1.0
      \\ \hline\hline
    \end{tabular}
  \end{center}
  \caption{ Comparison of solution methods for the multi-material
    problem. All methods were run with a stopping criterion of
    $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
    \sn{1}{-3}$.  The problem was run to an elapsed time of 100~ns.
    The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}
  \label{tab:multimat_comparison}
\end{table}
